Experimenting OpenMP/Python Coupling

### 2013-06-09 doc/papers/dls2013/paper.tex
Add dls paper

% to submit via https://www.easychair.org/account/signin.cgi?conf=europar2013
% max 12 pages
\documentclass{llncs}
%
\usepackage{listings}
\usepackage{url}
\usepackage{todo}
\usepackage{xcolor}
\newcommand\see{\emph{cf.\ }}

@@

\begin{document}
% update listing numbering
\renewcommand{\thelstlisting}{\arabic{lstlisting}}
%
\title{Pythran: Bringing OpenMP to a Python Subset}

\author{Serge Guelton\inst{1,2} \and Pierrick Brunet\inst{2}}

\institute{
    {\'E}cole Normale Sup{\'e}rieure, Paris, France
    \and
    T{\'e}l{\'e}com Bretagne, Plouzan{\'e}, France
}

@@

    The Python language offers an interesting solution as a scientific
    computing language thanks to the SciPy package. However, the lack of
    fine-grained parallelism support remains an important shortcoming. This
    paper proposes to add OpenMP directives to the Python language and shows
    that it is possible to turn a module written in a large Python subset
    enhanced with these directives into a native parallel module that runs as
    fast as its naive C equivalent, while keeping the input code totally
    compatible with the python interpreter. The approach is validated through
    the Pythran compiler.

\keywords{Python, OpenMP, C++, static compilation, scripting language.}
\end{abstract}

@@

This hybrid approach, based on a mix of interpreted and native code, is getting
widespread in the Python landscape. Section~\ref{sec:python-parallel} studies
the existing approaches and shows a critical lack of parallelism support, and
emphasises on the need for a backward-compatible approach.
Section~\ref{sec:python-openmp} studies the validity of using OpenMP directives
within the Python language in order to add fine-grain parallelism support to
Python in the context of a Python subset called Pythran. Static compilation of
this language into C++ and the underlying runtime is described in
Section~\ref{sec:python-static}. The Pythran compiler is benchmarked on several
scientific kernels and compared to Cython in Section~\ref{sec:validation}.

@@

In the many cores area, turning to parallelism to balance the performance
limitations of scripting languages, as described 
% in~\cite{choy05} in the context of the MATLAB language, or
in~\cite{mals07} in the context of the R language, seems legitimate. In the
context of Python, most approaches have focused on fork-based parallelism.

@@

@@

%There have been several approaches to replace the GIL by Transactional
%Memories~\cite{Riley2006,Tabba2010} but none of them made there way to the
%mainstream interpreters.

@@

Cython~\cite{cython2010} is such a Python dialect. It extends the syntax with
typing informations, calls to native functions from third party libraries, and a
limited set of parallelism constructs, such as the possibility to define
parallel loops, but no task parallelism. PLW~\cite{dongara2007} proposes a
similar approach that mixes Python with C, using raw strings to hold the C code
and the parallel directives, that are limited to parallel for loops.
%The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
%compiler uses additional type information to generate sequential LLVM bytecode.
PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
Python with kernels embedded as raw strings containing accelerator code.

@@

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
turn regular Python codes into native ones, but they do not offer support for
fine-grained parallelism beyond what the standard library proposes.
%~\footnote{Shedskin is compatible with several Python modules that provide
%coarse-grained, fork-based parallelism.}

@@

OpenMP is a standard API for parallel programming for Fortran, C and C++. It
consists in a set of parallelizing directives and a few runtime library calls.
If OpenMP is not activated, the directives are ignored, thus enabling
incremental parallelization of the original source code while keeping the
original code structure. The languages targeted by OpenMP are statically
compiled. This section studies the semantical adaptation required to use the
same API for a scripting language such as Python.

@@

While Pythran has a decorator mechanism\footnote{\see PEP 318,
\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
Python decorators.}, it only applies to functions, methods and classes and does
not make it possible to attach decorations to other statements. Pythran uses
the path taken by PLW~\cite{dongara2007}, that uses string instructions to hold
such decorations. As Python does not have anonymous block,
%~\footnote{\see PEP340, \url{http://www.python.org/dev/peps/pep-0340/} for a
%discussion concerning anonymous block support in Python.}
one has to create a dummy \texttt{if 1:} instruction to apply an annotation to a
whole block. Alternatively, the syntax \texttt{if 'my annotation':} is also
supported.

@@

Generator expressions are generators whose content is built from an other
iterator. For instance \texttt{(x*x for x in l)} successively yields the square
of each element in \texttt{l}. They behave like adaptors: they apply a
particular expression on each value of the iterator. It is also the case of the
\texttt{enumerate} builtin, that yields each element of the enumerated iterator
associated with its index. These generators are random iterators only if their
input iterator is a random iterator itself.

@@

Pythran runtime is aware of these three kind of iterators and accepts supports
parallel iteration over random access iterators and iterators adapting random
ones.

@@

\begin{figure}[ht]
    \caption{Relative speedups of Pythran+OpenMP version of the Hyantes prototype.}
    \label{fig:hyantes-speedup}
    \centering
    \includegraphics[width=\textwidth]{hyantes}
\end{figure}

@@

    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython implementation
    of a parallel reduction.}]
from libc.math cimport sqrt
from cython.parallel import parallel, prange
def sum_sqrt(double r):
    cdef int i
    with nogil, parallel():
        for i in prange(10000000): r += sqrt(i)
    return r
    \end{lstlisting}
%
    \begin{lstlisting}[label={lst:pythran-sample}, caption={Pythran implementation
    of a parallel reduction.}]
#pythran export sum_sqrt(float)
import math
def sum_sqrt(r):
    "omp parallel for private(i) reduction(+:r)"
    for i in xrange(10000000): r += math.sqrt(i)
    return r
    \end{lstlisting}
\end{figure}

@@

The source codes used for these benchmarks are available on the Pythran
repository.\footnote{\see \texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
branch.}

\begin{figure}[ht]
    \includegraphics[width=\textwidth]{cython}
    \caption{Comparison of Cython and Pythran generated code performance.}
    \label{fig:cython-pythran}
\end{figure}

@@

%=========================================================
\section*{Acknowledgments}
%=========================================================

@@

\bibliographystyle{splncs03}
\bibliography{biblio}

>>>>

% to submit via https://www.easychair.org/account/signin.cgi?conf=europar2013
% max 12 pages
\documentclass{sigplanconf}
%
\usepackage{listings}
\usepackage{url}
\usepackage{todo}
\usepackage{xcolor}
\newcommand\see{\emph{cf.\ }}

@@

\begin{document}
% update listing numbering
\renewcommand{\thelstlisting}{\arabic{lstlisting}}
%
\title{Experimenting OpenMP/Python Coupling}

\authorinfo{Serge Guelton}{{\'E}cole Normale Sup{\'e}rieure, Paris, France}{serge.guelton@telecom-bretagne.eu}
\authorinfo{Pierrick Brunet}{T{\'e}l{\'e}com Bretagne, Plouzan{\'e},France}{pierrick.brunet@telecom-bretagne.eu}

@@

    The Python language offers an interesting solution as a scientific
    computing language thanks to the SciPy package. However, the lack of
    fine-grained parallelism support remains an important shortcoming. This
    paper proposes to add OpenMP directives to the Python language and shows
    that it is possible to turn a module written in a large Python subset
    enhanced with these directives into a native parallel module that runs as
    fast as its naive C equivalent, while keeping the input code totally
    compatible with the Python interpreter. The approach is validated through
    the Pythran compiler.

\end{abstract}
\keywords{Python, OpenMP, C++, static compilation, scripting language.}

@@

This hybrid approach, based on a mix of interpreted and native code, is getting
widespread in the Python landscape. Section~\ref{sec:python-parallel} studies
the existing approaches and shows a critical lack of parallelism support, and
emphasises on the need for a backward-compatible approach.
Section~\ref{sec:python-openmp} studies the validity of using OpenMP directives
within the Python language in order to add fine-grain parallelism support to
Python in the context of a Python subset called Pythran~\cite{pythran2013}.
Static compilation of this language into C++ and the underlying runtime is
described in Section~\ref{sec:python-static}. The Pythran compiler is
benchmarked on several scientific kernels and compared to Cython in
Section~\ref{sec:validation}.

@@

In the many cores area, turning to parallelism to balance the performance
limitations of scripting languages, as described in~\cite{choy05} in the context
of the MATLAB language, or in~\cite{mals07} in the context of the R language,
seems legitimate. In the context of Python, most approaches have focused on
fork-based parallelism.

@@

There have also been several approaches to replace the GIL by Transactional
Memories~\cite{Riley2006,Tabba2010} but none of them made there way to the
mainstream interpreters.

@@

@@

Cython~\cite{cython2010} is such a Python dialect. It extends the syntax with
typing informations, calls to native functions from third party libraries, and a
limited set of parallelism constructs, such as the possibility to define
parallel loops, but no task parallelism. PLW~\cite{dongara2007} proposes a
similar approach that mixes Python with C, using raw strings to hold the C code
and the parallel directives, that are limited to parallel for loops.  The
numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}} compiler
uses additional type information to generate sequential LLVM bytecode.  PyCUDA
and PyOpenCL~\cite{klockner2012} also target accelerators by mixing Python with
kernels embedded as raw strings containing accelerator code.

@@

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
turn regular Python codes into native ones, but they do not offer support for
fine-grained parallelism beyond what the standard library proposes.

@@

OpenMP is a standard API for parallel programming for Fortran, C and C++. It
consists of a set of parallelizing directives and a few runtime library calls.
If OpenMP is not activated, the directives are ignored, thus enabling
incremental parallelization of the original source code while keeping the
original code structure. The languages targeted by OpenMP are statically
compiled. This section studies the semantical adaptation required to use the
same API for a scripting language such as Python.

@@

While Python has a decorator mechanism\footnote{\see PEP 318,
\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
Python decorators.}, it only applies to functions, methods and classes and does
not make it possible to attach decorations to other statements. Pythran uses the
path taken by PLW~\cite{dongara2007}, that uses string instructions to hold such
decorations. As Python does not have anonymous block,~\footnote{\see PEP340,
    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
anonymous block support in Python.} one has to create a dummy \texttt{if 1:}
instruction to apply an annotation to a whole block. Alternatively, the syntax
\texttt{if 'my annotation':} is also supported.

@@

Generator expressions are generators whose content is built from an other
iterator. For instance \texttt{(x*x for x in l)} successively yields the square
of each element in \texttt{l}. They behave like adaptors: they apply a
particular expression on each value of the iterator. It is also the case of the
\texttt{enumerate} builtin, that yields each element of the enumerated iterator
associated with its index. These generators are random access iterators only if their
input iterator is a random iterator itself.

@@

Pythran's runtime is aware of these three kind of iterators and supports
parallel iteration over random access iterators and iterators adapting random
ones.

@@

\begin{figure}[ht]
    \caption{Relative speedups of Pythran+OpenMP version of the Hyantes prototype.}
    \label{fig:hyantes-speedup}
    \centering
    \includegraphics[width=.5\textwidth]{hyantes}
\end{figure}

@@

    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython implementation
    of a parallel reduction.}]
from libc.math cimport sqrt
from cython.parallel import parallel, prange
def sum_sqrt(double r):
    cdef int i
    with nogil, parallel():
        for i in prange(10000000):
            r += sqrt(i)
    return r
    \end{lstlisting}
%
    \begin{lstlisting}[label={lst:pythran-sample}, caption={Pythran implementation
    of a parallel reduction.}]
#pythran export sum_sqrt(float)
import math
def sum_sqrt(r):
    "omp parallel for reduction(+:r)"
    for i in xrange(10000000):
        r += math.sqrt(i)
    return r
    \end{lstlisting}
\end{figure}

@@

The source codes used for these benchmarks are available on the Pythran
repository.\footnote{\see \texttt{https://github.com/serge-sans-paille/pythran}
    on \texttt{dls2013}
branch.}

\begin{figure}[ht]
    \includegraphics[width=.5\textwidth]{cython}
    \caption{Comparison of Cython and Pythran generated code performance.}
    \label{fig:cython-pythran}
\end{figure}

@@

%=========================================================
\acks
%=========================================================

@@

\bibliographystyle{abbrvnat}
\bibliography{biblio}
